Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2020-11-06 manuscript.rmd
1. SI figure number fixed, not dynamic anymore

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table \@ref(tab:param)), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space. In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

@@

To inversely estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq. \@ref(eq:M-chosen) and those in SI). Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions). We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein). $K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential. The use of maximum diurnal $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019]. To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see SI). Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.

@@

We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models. As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects the signature of source water. Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991]. We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig. S\@ref(fig:iso-data)). Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water deeper than 1 m, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.

@@

We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets [@Wolfe.2019; @Wolfe.2020] for six to seven species that overlapped with our assessment of ERD. We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins $\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6). See [SI](#SIData.brett) and Table S\@ref(tab:hyd-table) for data collection and estimation of these variables.

@@

The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)**a**, **b**), including the out-of-sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)**c**) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range (Table S\@ref(tab:param)) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs. S\@ref(fig:sens-sm-et), S\@ref(fig:sens-run)).

\
Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Hydrological droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig. S\@ref(fig:swp-droughts)). Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S\@ref(fig:swp-all-depths)).

@@

Parameters of leaf vulnerability curves were predictable from WSG and LMA (Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variance in parameter $B$ (Eq. \@ref(eq:b-mod), Adj. $R$^2^ = `r bfit.r2`, $p$ `r  bfit.p`). WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq. \@ref(eq:a-mod), *Adj. R*^2^ = `r afit.r2`, $p$ `r  afit.p`; also see Fig. S\@ref(fig:obs-plc-curves)). This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations (Table S\@ref(tab:ab-param), Fig. S\@ref(fig:obs-comm-plc-curves)).

@@

The best ERD model (with a multiplicative effect of VPD and not LAI, Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig. \@ref(fig:erd-iso)). Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<20 species vs. `r length(erd.sp)` species for the best model, Fig. S\@ref(fig:other-erd-mods)). For the rest of the analyses we therefore used ERD estimates from the less parsimonious, best model.

@@

Species with deeper ERD showed greater $K_\textrm{max,stem}$ (Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`, Fig. \@ref(fig:erd-hyd)**a**), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`, Fig. \@ref(fig:erd-hyd)**b**), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to xylem embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`, Fig. \@ref(fig:erd-hyd)**c**), and more negative above-ground hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~, $r$ = `r cor.hsm`, $p$ = `r p.hsm`, Fig. \@ref(fig:erd-hyd)**d**).

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. S\@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017]. Analysis of a species' realized hydraulic risk as the time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD experienced greater realized hydraulic risk and increasingly so over 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time)). For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic risk (Fig. \@ref(fig:erd-time)), but during the dry season ( Fig. \@ref(fig:swp-droughts)), and without elevated mortality rates (Fig. \@ref(fig:erd-mort-census-evg)). Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).

@@

Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs. \@ref(fig:sp-erd), \@ref(fig:erd-iso)). Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and more negative above-ground hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig. S\@ref(fig:erd-mort-census)), especially amongst the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time)). We thus demonstrate, for the first time, a link between realized drought risk, hydraulic architecture and long-term mortality outcomes for large tree species of a tropical forest.

@@

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for hydrological drought and trees' rooting depths. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig. \@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs. \@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census)). Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Figs. \@ref(fig:erd-mort-census-evg), \@ref(fig:erd-time)). Our results are consistent with studies that included three or four species each with known rooting or water-sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017]. Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010]. Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery of this important functional group [@Bennett:2015; @McGregor.2020]

\
Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought. The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see [@Goulden.2019; @Ivanov:2012]). In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts during 1990-2015. Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses. At BCI deep soil layers were recharged annually (Figs. S\@ref(fig:swp-all-depths), \@ref(fig:swp-droughts)), whereas in CT2018 they were not. The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.

@@

\
Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes. Future studies could better constrain the ERD model with higher frequency growth data such as those from dendrometer bands. At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015; @Mencuccini.2017], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models. Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

>>>>

To parameterize ELM-FATES, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site, else for the tropics (Table S1), and ran 5000 simulations using Latin Hypercube Sampling [LHS; @Stein.1987] from this global parameter space. In the Conrad catchment, in which the 50 ha plot is situated, soil hydraulic conductivity decreases with depth from 12.5 and 60 cm (@Godsey:2004; Table S1). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths <12.5 cm and >60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S1). For greater accuracy, instead of the ELM default pedo-transfer functions, we estimated parameters of soil retention curves using existing data for gravimetric water content (GWC) vs. $\mathrm{\Psi}$~soil~ for depths of 0.15, 0.4 and 1 m [@Kupers:2019]. We converted GWC to volumetric (VWC, $\theta$) using a measured bulk density value of 0.8 g cm^-3^ (see below). A @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences in water retention among depths were nominal,

@@

To inversely estimate effective rooting depth, we tested alternate empirical models for species-specific 5-yearly census growth, which use effective rooting depth as an input (Eq. \@ref(eq:M-chosen) and those in SI). Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ generally approaches $\mathrm{\Psi}$~soil~, we used $\mathrm{\Psi}$~soil~ in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which we assumed controls photosynthetic rate and growth (@Brodribb.2002; @Brodribb.2017; see Table \@ref(tab:symbols) for trait symbols and their definitions). We assumed $K_\textrm{leaf}$ to be a hydraulic bottleneck in the plant, and thus to determine or co-determine whole plant hydraulic response to soil drying; notably, root hydraulic vulnerability may also be a strong driver, and recent studies suggest that its vulnerability would be similar to, or correlated with that of leaves (see @Cochard.2020 and references therein). $K_\textrm{leaf}$ in our models, being based on vulnerability curves, is dynamic and thus depends on rooting depth and rhizosphere water potential. The use of maximum diurnal $\mathrm{\Psi}$~leaf~ based on pre-dawn $\mathrm{\Psi}$~soil~ as a proxy for diurnal $\mathrm{\Psi}$~leaf~ and its influence on growth at diurnal time-scales (to relate to 5-year average growth values) thus does not consider diurnal dynamics of $\mathrm{\Psi}$~leaf~ and other factors that would play a role in such diurnal dynamics, such as the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020; @Yang.2019]. To account for the non-linear impact of VPD on growth, we used predicted gross primary productivity (GPP; hereafter, $\hat{\textrm{VPD}}$) from a locally derived polynomial relationship between GPP and VPD (see Fig. S2 and SI). Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: $\hat{\textrm{VPD}}$, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Incorporation of radiation did not improve model fitting, possibly due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI.

@@

We used $\mathrm{\delta}$^2^H~xylem~ as an independent observation to validate the ERD models. As root water-uptake is generally a non-fractionating process, tree $\mathrm{\delta}$^2^H~xylem~ reflects the signature of source water. Given a vertical gradient of $\mathrm{\delta}$^2^H in soil and groundwater $\mathrm{\delta}$^2^H~xylem~ provides an index of rooting depth [@Dawson.1991]. We leveraged $\mathrm{\delta}$^2^H~xylem~ from BCI [@Meinzer:2001] for the dry-season of March 1997 as this period showed largest seasonal divergence in $\mathrm{\delta}$^2^H~xylem~ among species and vertically in soil and groundwater $\mathrm{\delta}$^2^H at natural abundance level (Fig. S3). Analyzing $\mathrm{\delta}$^2^H~xylem~ for 12 tree species @Meinzer:1999 found that four species sourced water deeper than 1 m, two other at least 0.8 m deep, one between 0 to 0.2 m, and five in the 0.2 to 0.7 m depth range over which the relationship between soil depth and $\mathrm{\delta}$^2^H~soil~ was variable.

@@

We evaluated whether ERD was associated with above-ground hydraulic traits sourcing the latter from a set of rare datasets [@Wolfe.2019; @Wolfe.2020] for six to seven species that overlapped with our assessment of ERD. We regressed species ERD against maximum stem hydraulic conductivity $K_\textrm{max,stem}$ (*n*=7), leaf turgor loss point $\mathrm{\Psi}$~tlp~ (*n*=7), vulnerability to embolism from cavitation $\mathrm{\Psi}$~88,stem~ measured in terms of pressure at which 88% of $K_\textrm{max,stem}$ is lost (*n*=7), and hydraulic safety margins $\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~ (*n*=6). See [SI](#SIData.brett) and Table S2 for data collection and estimation of these variables.

@@

The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)**a**, **b**), including the out-of-sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)**c**) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was sensitive to the Ball-Berry stomatal slope parameter, the ELM root distribution parameter that regulates the depth of the rooting profile, soil hydraulic conductivity profile especially at depth, and the adjustment factor that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs. S4, S5).

\
Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had an extreme hydrological drought year, but census intervals varied in terms of number of extreme years, drought intensity at depth, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Hydrological droughts in census intervals 2000-2005 and 2005-2010 distinctly occurred in the wet season, with one to two extreme hydrological drought years each, depending on the depth location (Fig. \@ref(fig:swp-droughts)). Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S6).

@@

Parameters of leaf vulnerability curves were predictable from WSG and LMA (Fig. \@ref(fig:ab-fit)). WSG and LMA explained a large proportion of variance in parameter $B$ (Eq. \@ref(eq:b-mod), Adj. $R$^2^ = `r bfit.r2`, $p$ `r  bfit.p`). WSG and parameter $B$ explained a large proportion of variance in parameter $A$ (Eq. \@ref(eq:a-mod), *Adj. R*^2^ = `r afit.r2`, $p$ `r  afit.p`; also see Fig. S7). This predictive power allowed us to estimate leaf vulnerability curves used in the ERD models for `r sp.n.model` out of `r length(erd.sp)` ERD species that lacked direct observations (Table S3, Fig. S8).

@@

The best ERD model (with a multiplicative effect of VPD and not LAI, Eq. \@ref(eq:M-chosen)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig. \@ref(fig:erd-iso)). Two of the alternative models with LAI had a strong relationship with $\mathrm{\delta}$^2^H~xylem~ as well, but one that was based on a smaller set of species ($n$ = 4), because fewer species passed the criteria of sufficient species-level goodness-of-fit for these models (<20 species vs. `r length(erd.sp)` species for the best model, Fig. S9). For the rest of the analyses we therefore used ERD estimates from the less parsimonious, best model.

@@

Species with deeper ERD showed greater $K_\textrm{max,stem}$ (Spearman's' $r$ = `r cor.k`, $p$ = `r p.k`, Fig. \@ref(fig:erd-hyd)**a**), less negative leaf $\mathrm{\Psi}$~tlp~ ($r$ = `r cor.tlp`, $p$ = `r p.tlp`, Fig. \@ref(fig:erd-hyd)**b**), less negative $\mathrm{\Psi}$~88,stem~ and thus greater vulnerability to xylem embolism from cavitation ($r$ = `r cor.stem.88`, $p$ = `r p.stem.88`, Fig. \@ref(fig:erd-hyd)**c**), and more negative above-ground hydraulic safety margins ($\mathrm{\Psi}$~min~ - $\mathrm{\Psi}$~88,stem~, $r$ = `r cor.hsm`, $p$ = `r p.hsm`, Fig. \@ref(fig:erd-hyd)**d**). See Fig. S10 for the full correlation matrix.

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species mortality rates significantly decreased with deeper ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but still significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. S11). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). While our $\mathrm{\Psi}_{\textrm{soil}}$ estimates did not extend prior to 1990, due to a lack of reliable climate forcing, a prior study found that during the period 1985-1990, for which ERD explained significant mortality, BCI experienced three extreme deficit water years, following the 1983 El-Niño year [@Condit.2017]. Analysis of a species' realized hydraulic risk as the time spent under significant hydrological drought (% days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD experienced greater realized hydraulic risk and increasingly so over 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time)). For the census interval 1990-1995, species with the shallowest ERD also showed greater realized hydraulic risk (Fig. \@ref(fig:erd-time)), but during the dry season ( Fig. \@ref(fig:swp-droughts)), and without elevated mortality rates (Fig. \@ref(fig:erd-mort-census-evg)). Overall, species with ERD deeper than 1.7 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa).

@@

Our novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil moisture profiles, and observed VPD was consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs. \@ref(fig:sp-erd), \@ref(fig:erd-iso)). Our analyses revealed that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but not safety (Fig. \@ref(fig:erd-hyd)). Species with deeper ERD showed higher maximum stem hydraulic conductance, and greater xylem vulnerability to embolism and more negative above-ground hydraulic safety margins compared to shallow ERD species. Nevertheless, deeper ERD species showed lower mortality rates than shallower ERD species in significant wet season droughts over the 35-yr record (1981-2015) (Fig. S11), especially amongst the evergreen species (Fig. \@ref(fig:erd-mort-census-evg)). Analysis of time spent beyond $\mathrm{\Psi}$~crit~ in the soil zone matching ERD $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ showed that this was likely due to decreasing realized risk of hydraulic failure with increasing access to water for species with deeper ERD (Fig. \@ref(fig:erd-time)). We thus demonstrate, for the first time, a link between realized drought risk, hydraulic architecture and long-term mortality outcomes for large tree species of a tropical forest.

@@

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for hydrological drought and trees' rooting depths. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig. \@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk [@Anderegg:2016], but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Figs. \@ref(fig:swp-droughts), \@ref(fig:erd-mort-census-evg), S11). Deeper ERD species on average had more hydraulically efficient, but risky, suite of above-ground traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Figs. \@ref(fig:erd-mort-census-evg), \@ref(fig:erd-time)). Our results are consistent with studies that included three or four species each with known rooting or water-sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017]. Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010]. Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery of this important functional group [@Bennett:2015; @McGregor.2020]

\
Our result--of lower mortality in deep-rooted trees--contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally tropical dry forest had higher mortality in a rare, prolonged drought. The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see [@Goulden.2019; @Ivanov:2012]). In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts during 1990-2015. Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses. At BCI deep soil layers were recharged annually (Figs. S6, \@ref(fig:swp-droughts)), whereas in CT2018 they were not. The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.

@@

\
Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes. Future studies could better constrain the ERD model with higher frequency growth data such as those from dendrometer bands. At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015; @Mencuccini.2017], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models. Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig. S9). We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].
